Shallow EDSLs and Object-Oriented Programming

### 2016-06-16 paper/sections/Introduction.tex
more text

This definition seems to be widely accepted and similar definitions
appear in many works~\cite{}. We believe that this definition is a
vague and not very accurate, often leading to some contradicting
claims. For example, in their work on EDSLs~\cite{}, Gibbons and Wu
claim that deep embeddings (which encode ASTs using algebraic
datatypes in Haskell) allow adding new DSL interpretations easily, but
they make adding new language constructs difficult. In contrast
Gibbons and Wu claim that shallow embeddings have dual modularity
properties: new cases are easy to add, but new interpretations are
hard.  A contradiction happens if instead of using Haskell and
algebraic datatypes, one uses an OO language to encode an AST, for
example with the {\sc Composite} pattern. In that case, according to
definition of deep embedding above, the OO approach classifies as a
deep embedding. Nevertheless, as the Expression Problem~\cite{} tell
us, in the OO approach adding new cases is easy, but adding
interpretations is hard. Thus we have a contradiction, since we have
an AST representation (i.e. a deep embedding) with the modularity
properties of shallow embeddings.

We believe that the core of problem is that ASTs can be represented 
in multiple ways. In particular, it is well know that functions alone 
are enough to encode datastructures such as ASTs (via Church
encodings~\cite{}). Distinguishing deep and shallow
embeddings solely on whether a datastructure is being used 
or whether functions are used to encode interpretations directly 
falls short when the functions themselfves represent an AST.

In this paper we propose a different definition for deep and shallow
embeddings, by distinguishing on what kind of data abstraction is used
by each encoding. Deep embeddings use user-defined types
(i.e. disjoint unions or algebraic datatypes), whereas shallow
embeddings use procedural abstraction. This distinction is inspired by
Reynolds 1975 paper~\cite{} on user-defined types and procedural
abstraction being complementary approaches to data abstraction, which
many view~\cite{} as the first documented discussion on the Expression
Problem.

>>>>

This definition seems to be widely accepted and similar definitions
appear in many works~\cite{}. We believe that this definition is
vague, and often leads to some contradicting
claims. For example, in their work on EDSLs~\cite{}, Gibbons and Wu
claim that deep embeddings (which encode ASTs using algebraic
datatypes in Haskell) allow adding new DSL interpretations easily, but
they make adding new language constructs difficult. In contrast
Gibbons and Wu claim that shallow embeddings have dual modularity
properties: new cases are easy to add, but new interpretations are
hard.  A contradiction happens if instead of using Haskell and
algebraic datatypes, one uses an OO language to encode an AST, for
example with the {\sc Composite} pattern. In that case, according to
definition of deep embedding above, the OO approach classifies as a
deep embedding. Nevertheless, as the Expression Problem~\cite{} tell
us, in the OO approach adding new cases is easy, but adding
interpretations is hard. Thus we have a contradiction, since we have
an AST representation (i.e. a deep embedding) with the modularity
properties of shallow embeddings.

We believe that the core of problem is that ASTs can be represented in
multiple ways. In particular, it is well know that functions alone are
enough to encode datastructures such as ASTs (via Church
encodings~\cite{}). Distinguishing deep and shallow embeddings based
solely on whether a datastructure is being used or whether
interpretation functions are used is misleading. Moreover, it gives
the false impression that shallow embeddings are somewhat less
expressive than deep embeddings, because they do not have access to
the AST.

In this paper we propose to distinguish EDSLs in terms of the data
abstraction used to model the language constructs instead.  We follow
Reynold's classification~\cite{} of data abstractions:
\emph{procedural abstraction} and \emph{user-defined types}. It is
clear that shallow embeddings use \emph{procedural
  abstraction}~\cite{}: the DSLs are modelled by interpretation
functions. Therefore, the other implementation option for EDSLs is to use
\emph{user-defined types}. In Reynolds terminology user-defined types
mean disjoint union types, which are nowadays commonly available in
modern languages as \emph{algebraic datatypes}. Disjoint union types 
can also be emulated in OOP using the {\sc Visitor} pattern.

As Cook~\cite{} argued
procedural abstraction is the essence of Object-Oriented
Programming. If we accept Cook's view, the implementation of a shallow
DSL in OOP languages should simply correspond to a standard
object-oriented program. By identifying shallow EDSLs with procedural
abstraction it is clear that an EDSL implemented in a functional
language using an interpretation function is closely related to an
Object-Oriented Program. The new interpretation of shallow EDSL
removes the contradiction in Gibbons and Wu's work. Both the Haskell
implementation of a shallow EDSL and the OOP program using the {\sc
  Composite} pattern are now under the definition of shallow
EDSL. Following Reynold's work on data abstraction~\cite{}, the other
implementation option for EDSLs is to use \emph{user-defined
  types}. In his original work user-defined types mean disjoint union
types, which are nowadays commonly available in modern languages as
\emph{algebraic datatypes}.

By using the kind of data abstraction as a way to distinguish 
between different implementations strategies for DSLs it is clear 
that the OOP program and a correponding

Instead of shallow
EDSLs, we prefer to talk about \emph{procedural} EDSLs. The new
terminology is important for two reasons: 1) it is more precise; 
2) procedural EDSLs emcompass a wider range

 not all procedural EDSLs

%However it is this, apparently harmless, translation step is
%where the distinction between shallow and deep embeddings becomes
%fuzzy, since it can be argued that the OO program represents an AST.

Instead of deep embeddings, we prefer

propose a different definition for deep and shallow
embeddings, by distinguishing on what kind of data abstraction is used
by each encoding. Deep embeddings use user-defined types
(i.e. disjoint unions or algebraic datatypes), whereas shallow
embeddings use procedural abstraction. This distinction is inspired by
Reynolds 1975 paper~\cite{} on user-defined types and procedural
abstraction being complementary approaches to data abstraction, which
many view~\cite{} as the first documented discussion on the Expression
Problem.
